ASSIGNMENT 1 - LENDING CLUB LOANS DATA
========================================================
Load required libraries
```{r}
library(lsr)
```

Read in the data and check the structure of the data along with the first few rows
```{r}
data <- read.csv('C:\\Users\\Bahae Omid\\Desktop\\Data Analysis with R\\Assignments\\Assignment 1\\Data\\Raw Data\\loansdata.csv')
str(data)
head(data)
```

Data Cleaning:coerce interest rate, FICO score, Employment Length and Debt to Income Ratio to numeric
```{r}
data$Interest.Rate <- as.numeric(gsub('[%]','',data$Interest.Rate))
data$Debt.To.Income.Ratio <- as.numeric(gsub('[%]','',data$Debt.To.Income.Ratio))
#Replace FICO Range with upper value of the range
data$FICO.Range <- as.numeric(gsub('[0-9]+-','',data$FICO.Range))
#Replacing employment length of <1 year with 0, 10+ years with 10, and n/a with mean of the column
data[data[,14]=='< 1 year',14]<-0
data[data[,14]=='10+ years',14]<-10
data$Employment.Length <- gsub(' years| year','',data$Employment.Length)
data$Employment.Length <- as.numeric(data$Employment.Length)
data[is.na(data$Employment.Length),14] <- mean(data$Employment.Length,na.rm=T)
```

Exploratory Analysis: Data summary
```{r}
summary(data)
```

Although there's an unusual data point in amount funded by investors(-0.01) but we're going to ignore this since we won't need this column for our analysis
```{r}
data <- data[-2]
```

There's a very odd data point in monthly income($102,750). We're going to assume this to be an outlier since it's very unlikely for somebody with this monthly income to ask for a small loan
```{r}
data <- data[-which.max(data$Monthly.Income),]
```

Scanning for missing values
```{r}
sapply(data,function(x) sum(is.na(x)))
#Check the rows with missing values and see if they can be excluded from the analysis
data[which(is.na(data$Open.CREDIT.Lines)),]
#Removing the rows with missing values since they won't affect our analysis significantly
data <- data[-which(is.na(data$Open.CREDIT.Lines)),]
```

Checking correlations among all variables and plotting pairwise scatter for illustration
```{r}
cor(data[,sapply(data,class)!='factor'])
pairs(data[,sapply(data,class)!='factor'])
```
A very strong negative correlation between Interest rate and FICO Score can be seen. Also, amount requested seems to have some association with Interest rate. Therefore, these two variables could be potential candidates to explain the dependant variable. 
We're going to exclude Revolving Credit Balance, Open Credit Lines, and Debt to Income Ratio as these seem to be highly correlated(not from the correlation matrix necessarily, rather from domain knowledge) because these variables are already confounded within the FICO Score(a research on the way the credit score is calculated reveals the fact). so we can conclude that these variables are reduntant. Including reduntant variables may skew the model due to multicollinearity.

We can do a quick check through running a linear regression model to prove this
```{r}
summary(lm(data$Interest.Rate~data$Revolving.CREDIT.Balance+data$Open.CREDIT.Lines+data$Debt.To.Income.Ratio))
```
It's clear that this poor model (R2=0.03) contain variables that do not explain the interest rate.

Domain knowledge allows us to omit Inquiries in the past 6 months along with monthly income and employment length as these effects are already reflected in FICO score. so they are considered to be reduntant.
```{r}
data <- data[-c(8,12,13)]
```

So far, we've looked at the effect of the numeric variables on Interest rate. Let's generate a few boxplots to examine the effect of the factor variables on Interest Rate
```{r fig.width=7, fig.height=6}
plot(data$Interest.Rate~data$Loan.Length)
plot(data$Interest.Rate~data$Loan.Purpose)
plot(data$Interest.Rate~data$State)
plot(data$Interest.Rate~data$Home.Ownership)
```
Except for Home Ownership, the graphs suggest that there may be a difference in the mean of the groups.

let's do analysis of variance to see if this is the case (check the assumptions of anova before using this technique)
```{r}
model.test <- aov(data$Interest.Rate~data$State+data$Loan.Length+data$Loan.Purpose)
etaSquared(model.test)
```
from the R2 values we can see that only loan length can potentially imporve our regression model in explaining the Interest Rate

After eliminating all the redundant and insignificant variables, we are ready to construct our final model with FICO Score, Amoun Requested, and Loan Length being the independant variables
```{r}
model.final <- lm(data$Interest.Rate~data$FICO.Range+data$Amount.Requested+data$Loan.Length)
summary(model.final)
```
It's clreat that all variables are statistically significant and the model is vrey strong (R2=0.75) explaining about 75% of the variation in Interest Rate

A plot of the final model to illustrate the associations
```{r, fig.width=10, fig.height=6}
par(mfrow=c(1,2))
plot(data$Interest.Rate~data$FICO.Range, col=data$Loan.Length, pch=19,xlab='FICO Score',ylab='Interest Rate')
abline(c(lm(data$Interest.Rate~data$FICO.Range+data$Loan.Length)$coefficients[1],lm(data$Interest.Rate~data$FICO.Range+data$Loan.Length)$coefficients[2]),col='black',lwd=3)
abline(c(lm(data$Interest.Rate~data$FICO.Range+data$Loan.Length)$coefficients[1]+lm(data$Interest.Rate~data$FICO.Range+data$Loan.Length)$coefficients[3],lm(data$Interest.Rate~data$FICO.Range+data$Loan.Length)$coefficients[2]),col='red',lwd=3)
legend('topright',legend=unique(data$Loan.Length),col=c('black','red'),pch=c(19,19),cex=0.75 )
mtext(text='(a)',side=3,line=1)
plot(data$Interest.Rate~data$Amount.Requested, col=data$Loan.Length, pch=19,xlab='Loan Amount Requested',ylab='Interest Rate')
abline(c(lm(data$Interest.Rate~data$Amount.Requested+data$Loan.Length)$coefficients[1],lm(data$Interest.Rate~data$Amount.Requested+data$Loan.Length)$coefficients[2]),col='black',lwd=3)
abline(c(lm(data$Interest.Rate~data$Amount.Requested+data$Loan.Length)$coefficients[1]+lm(data$Interest.Rate~data$Amount.Requested+data$Loan.Length)$coefficients[3],lm(data$Interest.Rate~data$Amount.Requested+data$Loan.Length)$coefficients[2]),col='red',lwd=3)
legend('topleft',legend=unique(data$Loan.Length),col=c('black','red'),pch=c(19,19),cex=0.75)
mtext(text='(b)',side=3,line=1)
par(mfrow=c(1,1))
```
As seen in figure (a), FICO Score is highly associated with Interest Rate and higher interest rates correspond to longer-term loans and lower FICO scores. figure (b) shows a strong association between Loan Amount Requested and Interest Rate and as seen, higher interest rates pertain to longer-term loans and higher amount requested.

Regression Diagnostics: checking for validity of regression model assumptions
```{r}
#Storing the residulas of the regression model and plotting the residuals
model.res <- resid(model.final)
plot(model.res,pch=19)
abline(h=mean(model.res),col='red',lwd=2)
```{r, fig.width=14, fig.height=6}
#Checking for correlations between residuals and the independant variables
par(mfrow=c(1,3))
plot(data$FICO.Range,model.res,col=data$Loan.Length, pch=19)
plot(data$Amount.Requested,model.res,col=data$Loan.Length, pch=19)
plot(data$Interest.Rate,model.res,col=data$Loan.Length,pch=19)
par(mfrow=c(1,1))
```
The model residuals seem to be random with average of near 0 and constant variation. Moreover, the residuals do not correlate with the independant and dependant variables. so the assumption of normality is met.